---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# parquetMS

<!-- badges: start -->

<!-- badges: end -->

ParquetMS can convert HDMSe data acquired with Unifi to tabular format for use in R or Python, with a small filesize when saved on disk.

Two output data file formats can be obtained:

-   the [Apache Parquet](https://parquet.apache.org/) format for minimal filesize and fast access. Two files are produced: one for MS data, one for metadata.

-   the [HDF5](https://www.hdfgroup.org/solutions/hdf5/) format with all data and metadata in one file, fast access but larger filesize.

## Installation

You can install parquetMS in R with the following command:

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("leesulab/parquetMS")
```

To use the HDF5 format, the `rhdf5` package needs to be installed:

```{r eval=FALSE}
install.packages("BiocManager")
BiocManager::install("rhdf5")
```

## Usage

First load the package:

```{r eval=FALSE}
library("parquetMS")
```

```{r include=FALSE}
library("parquetMS")
```

Then create connection parameters to the Unifi API (retrieve token).

```{r eval=FALSE}
con = create_connection_params(apihosturl = "http://localhost:50034/unifi/v1", identityurl = "http://localhost:50333/identity/connect/token")
```

If the `parquetMS` and `R` session are run from another computer than where the Unifi API is installed, replace `localhost` by the IP address of the Unifi API.

```{r eval=FALSE}
con = create_connection_params(apihosturl = "http://192.0.2.0:50034/unifi/v1", identityurl = "http://192.0.2.0:50333/identity/connect/token")
```

```{r include=FALSE}
#con = create_connection_params(apihosturl = "http://10.12.3.154:50034/unifi/v1", identityurl = "http://10.12.3.154:50333/identity/connect/token" )

con = create_connection_params(apihosturl = "http://localhost:50034/unifi/v1", identityurl = "http://localhost:50333/identity/connect/token")
```

Now we can use these connection parameters to access the Unifi folders. The following function will show the list of folders and their IDs (e.g. `abe9c297-821e-4152-854a-17c73c9ff68c` in the example below).

```{r eval=FALSE}
folders = folders_search(con)
folders
```

```{r echo=FALSE}
folders = folders_search(con)
folders[3:4,]
```

With a folder ID, we can access the list of Analysis items in the folder:

```{r eval=FALSE}
ana = analysis_search(con, "abe9c297-821e-4152-854a-17c73c9ff68c")
ana
```

```{r eval=FALSE, include=FALSE}
ana = analysis_search(con, "abe9c297-821e-4152-854a-17c73c9ff68c")
ana[4:6,]
```

Finally, with an Analysis ID, we can get the list of samples (injections) acquired in this Analysis:

```{r eval=FALSE}
samples = sample_search(con, "e236bf99-31cd-44ae-a4e7-74915697df65")
samples
```

```{r eval=FALSE, include=FALSE}
samples = sample_search(con, "e236bf99-31cd-44ae-a4e7-74915697df65")
samples[2:5,]
```

Once we get a sample ID, we can use it to download the sample data:

```{r eval=FALSE}
convert_one_sample_data(con, sample_id = "0134efbf-c75a-411b-842a-4f35e2b76347", "sample1", "analysis_name")
```

This command will create a folder named `analysis_name` in the working directory and save the sample data with the name `sample1.parquet` and its metadata with the name `sample1metadata.parquet`.

With an Analysis ID, we can convert and save all samples from the chosen Analysis:

```{r eval=FALSE}
convert_all_samples_data(con, analysis_id = "e236bf99-31cd-44ae-a4e7-74915697df65", "analysis_name")
```

To use the HDF5 format instead of Parquet, the format argument can be used as below:

```{r eval=FALSE}
convert_one_sample_data(con, sample_id = "0134efbf-c75a-411b-842a-4f35e2b76347", "sample1", "analysis_name", format = "hdf5")

convert_all_samples_data(con, analysis_id = "e236bf99-31cd-44ae-a4e7-74915697df65", "analysis_name", format = "hdf5")
```

This will save the samples data and metadata in the same `file.h5` file.

Parquet or HDF5 files can be opened easily in `R` with the `arrow` or `rhdf5` packages. Parquet files contain both low and high energy spectra (HDMSe), and HDF5 files contain low energy in the "ms1" dataset, high energy in the "ms2" dataset, and metadata in the "metadata" dataset.

```{r eval=FALSE}
sampleparquet = arrow::read_parquet("sample1.parquet")
metadataparquet = arrow::read_parquet("sample1metadata.parquet")

samplehdf5 = rhdf5::h5read("sample1.h5", name = "ms1")
samplehdf5 = rhdf5::h5read("sample1.h5", name = "ms2")
samplehdf5 = rhdf5::h5read("sample1.h5", name = "metadata")
```
