---
title: "UNIFI API configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{UNIFI API configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(arcMS)
```

# UNIFI API configuration

The UNIFI API must be configured properly for `arcMS` to connect correctly.

The installation, configuration and connection parameters all depend on the version of UNIFI running. It is essentially the same configuration than for using Proteowizard msconvert software for conversion.

## With UNIFI

Details about the Waters UNIFI API installation and configuration can be found in <https://www.waters.com/webassets/cms/support/docs/715005443rc.pdf>

### Installation

The UNIFI Web API is not directly installed with a UNIFI Workstation, it must be installed from the installation DVD drive or the folder where the installation files were copied, with the `UNIFIWebAPISetup.exe` installer.

### Configuration

Applications to be used with the Web API must be registered in the UNIFI Client Authorization Application. It can be accessed from the UNIFI Administration page, by clicking "External Application Security Configuration". It will open a browser with a page to connect to the Waters UNIFI Identity Server (at the URL `localname:50333/identity/admin#/`). Users must connect (Login button) with their credentials (typically `administrator`/`administrator`).

A "new client" must be created with the following configuration:

|                |                     |
|----------------|---------------------|
| Client Name    | resourceownerclient |
| Client Id      | resourceownerclient |
| OAuth 2.0 Flow | ResourceOwner       |

Then it must be enabled and edited with the following parameters:

|                       |        |
|-----------------------|--------|
| Client Secret         | secret |
| Access token lifetime | 86400  |
| Scope                 | unifi  |

The scope should have been created automatically but the secret value must be added manually.

### Connection parameters

The URL to be used for connection and retrieval of token are the following (if not changed during installation):

```{r, eval=FALSE}
con = create_connection_params(apihosturl = "http://localhost:50034/unifi/v1", identityurl = "http://localhost:50333/identity/connect/token")
```

Default `username` and `password` are `administrator` and `administrator` but you can pass other values to the command above if necessary.

## With waters_connect UNIFI

Details about the waters_connect configuration can be found in this video: <https://videos.waters.com/detail/video/6342765707112/set-up-msconvert-with-waters_connect-3.0.0>

### Installation

With waters_connect, the API is directly installed, no additional installation step is required.

### Configuration

Applications to be used with the Web API must be registered with the ApplicationRegistrationTool.exe app (typically installed in the `C:\Program Files\Waters\waters_connect\ApplicationRegistrationTool\` folder). It is a terminal application that must be executed with admin rights (right click, Run as administrator). Users must connect with their credentials (typically `administrator`/`administrator`).

A new client application can be registered by typing R, followed by the following configuration:

|                         |                             |
|-------------------------|-----------------------------|
| Application client name | resourceownerclient         |
| Flow                    | resource owner password (R) |
| Access token type       | Reference token (R)         |
| Secret                  | secret                      |

Client scope is automatically set as `webapi`.

### Connection parameters

The URL to be used for connection and retrieval of token are the following:

```{r, eval=FALSE}
con = create_connection_params(apihosturl = "https://localhost:48505/unifi/v1", identityurl = "https://localhost:48333/connect/token")
```

Default `username` and `password` are `administrator` and `administrator` but you can pass other values to the command above if necessary.
