---
title: "Distant query to Parquet file"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distant query to Parquet file}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A Parquet file can be hosted online and queried almost with the same methods shown in the `vignette("data-filtration-tutorial")` vignette.

We can use DuckDB again, but need to create a database distant connection first before we can make our query in the dplyr syntax.

```{r eval=FALSE}
library(DBI)
library(dplyr)
library(duckdb)
```

```{r eval=FALSE}
con = dbConnect(duckdb::duckdb(), dbdir=":memory:", read_only=FALSE)

dbExecute(con, "INSTALL httpfs;") # run once to install the httpfs library
dbExecute(con, "LOAD httpfs;")

dbExecute(con,
  "CREATE VIEW cal1 AS
   SELECT * FROM PARQUET_SCAN('https://github.com/leesulab/arcms-dataviz/raw/refs/heads/main/cal1-full-long.parquet');
")
```

Check that the connection has been created:
```{r eval=FALSE}
dbListTables(con)
```

Then make the query:

```{r eval=FALSE}
tic = tbl(con, "cal1") |>
  filter(mslevel == 1) |>
  group_by(rt) |> 
  summarise(intensity = sum(intensity)) |> 
      arrange(rt) |>
  collect()
```  

The TIC data has been downloaded and can be plotted as shown in other vignettes:

```{r eval=FALSE}
plotly::plot_ly(tic, 
                y=~intensity, 
                x=~rt, 
                type = 'scatter',
                mode = 'lines',
                line = list(color = 'rgba(0,0,0,1)', width = 1),
                line = list(shape = 'spline', smoothing = 1)
)
```


