---
title: "Open Parquet files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Open Parquet files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(arcMS)
```

## How to open converted files?

Parquet files created by `arcMS` can be easily opened with R or Python. There are also functions included in the `arcMS` package to recreate `sample_dataset` objects from Parquet files. This class of object is associated with generic functions to quickly get data, metadata and simple plots like TIC.

### 1. Load converted Parquet files in R without arcMS

Files created by `arcMS` in the Parquet format can be opened in R with the Arrow library.

Data can be loaded in RAM with the following commands:

```{r, include = FALSE}
#library(arrow)
#download.file("https://media.githubusercontent.com/media/leesulab/arcms-dataviz/main/cal1-full-long.parquet", "cal1.parquet")
# data = read_parquet("cal1.parquet")
data = readRDS("hdata.rds")
```

```{r, eval = FALSE}
library(arrow)
data = read_parquet("converted_file.parquet")
```

```{r}
head(data)
```

Data can then be filtered and quickly aggregated, e.g. to obtain TIC plot:

```{r, eval = FALSE}
library(data.table)
data = as.data.table(data)
datalow = data[mslevel == 1, ]
TIC = datalow[, list(intensity = sum(intensity)), by=list(rt)]
plot(TIC$rt, TIC$intensity, type = "l")
```

```{r echo=FALSE}
TIC = readRDS("tic.rds")
plot(TIC$rt, TIC$intensity, type = "l")
```

To save RAM, data can also be manipulated directly on-disk thanks to the Arrow library (`open_dataset()` function):

```{r, eval = FALSE}
data = open_dataset("converted_file.parquet")
```

Data can be filtered, rearranged, sorted and aggregated with the dplyr syntax, and only the resulting data will be loaded in RAM (with the `collect()` function):

```{r, eval = FALSE, message=FALSE}
library(dplyr)

TIC = data |>
  filter(mslevel == "1") |>
  arrange(rt) |>
  group_by(rt) |> 
  summarise(int = sum(intensity)) |> 
  collect()
```

### 2. Load converted Parquet files in Python

Parquet files can be opened as a DataFrame in Python with the pandas library:

```{python, include = FALSE, eval = FALSE, python.reticulate = FALSE}
import pandas as pd
data = pd.read_parquet("cal1.parquet")
```

```{python, eval = FALSE, python.reticulate = FALSE}
import pandas as pd
data = pd.read_parquet("converted_file.parquet")
ms1 = data[data['mslevel'] == "1"]
print(ms1)
```

It can also be loaded as an Arrow object (ParquetDataset) with the pyarrow library:

```{python, include = FALSE, eval = FALSE, python.reticulate = FALSE}
import pyarrow.parquet as pq
data = pq.ParquetDataset("cal1.parquet")
```

```{python, eval = FALSE, python.reticulate = FALSE}
import pyarrow.parquet as pq
data = pq.ParquetDataset("converted_file.parquet")
```

### 3. Load converted Parquet files in R with arcMS

The methods above only retrieve the main data from the Parquet file, but not metadata. Other functions must be used to retrieve metadata.

To simplify opening both data and metadata, functions are available in `arcMS` to load a Parquet file as a `sample_dataset` object, also allowing easy manipulation with some generic functions.

```{r, eval= FALSE}
library(arcMS)
dataset = create_sample_dataset("converted_file.parquet")  
```

Retrieve main data:

```{r, eval = FALSE}
data = get_sample_data(dataset)
```

```{r}
head(data)
```

Retrieve sample metadata:

```{r, eval = FALSE}
sample_metadata = get_sample_metadata(dataset)
```

```{r, include = FALSE}
sample_metadata = readRDS("sample_metadata.rds")
```
```{r}
str(sample_metadata)
```

Retrieve spectrum metadata: 

```{r, eval = FALSE}
spectrum_metadata = get_spectrum_metadata(dataset)
```

```{r, include = FALSE}
spectrum_metadata = readRDS("spectrum_metadata.rds")
```
```{r}
str(spectrum_metadata)
```

